---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# confcons

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

TODO: A paragraph that describes the high-level purpose of the package.

## Installation

You can install the development version of confcons from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github(repo = "bfakos/confcons", upgrade = "never")
```

TODO: install with vignette

## Examples

TODO: An example that shows how to use the package to solve a simple problem.
TODO: An overview that describes the main components of the package. For more complex packages, this will point to vignettes for more details.

```{r example}
library(confcons)
# TODO
```

